# Binomial Test

## Binomial Test

The binomial test is used to look for evidence that the proportion of a Binomial distributed random variable may differ from a hypothesized (or previously observed) value.

### Test Statistic

The test statistic for a binomial test is the observed frequency of experimental subjects that exhibit the trait of interest.  

### Definitions

Let $X$ be a random variable following a binomial distribution with parameters $n$ and $\pi$. Let $x$ be the observed frequency of experimental subjects exhibiting the trait of interest. 

### Hypotheses

The hypotheses for the Binomial test may take the following forms:

For a two-sided test:

$$\begin{align}
H_0: \pi = \pi_0 \\
H_a: \pi \neq \pi_0
\end{align}$$

For a one-sided test:

$$\begin{align}
H_0: \pi < \pi_0 \\
H_a: \pi \geq \pi_0
\end{align}$$

or 

$$\begin{align}
H_0: \pi > \pi_0 \\
H_a: \pi \leq \pi_0
\end{align}$$

### Decision Rule

The decision to reject the null hypothesis is made when the observed value of $x$ lies in the critical region that suggests the probability of that observation is low. We define the critical region as the upper bound we are willing to accept for $\alpha$, the Type I Error.

In a two-sided test, the upper bound is shared equally in both tails. Due to the discrete nature of the distribution, the total probability in the tails may not equal $\alpha$. The figures below depict examples of rejection regions for selected values of the Binomial distribution parameters. The decision rule is:

Reject $H_0$ if $x < Binomial(\alpha/2, n, \pi_0)$ or $x > Binomial(1 - \alpha/2, n, \pi_0)$

```{r, echo = FALSE, fig.path = 'figures/', fig.cap='The examples displayed use $n = 20$. For the top, middle, and bottom examples, $\\pi$ is set at 0.3, 0.5, and 0.75, respectively. Notice that in some cases, the rejection regions for $\\alpha = 0.10$ and $\\alpha = 0.05$ are identical.'}
DF <- 
  bind_rows(
    data.frame(x = 0:20,
               n = 20,
               p = .3),
    data.frame(x = 0:20,
               n = 20,
               p = .5),
    data.frame(x = 0:20,
               n = 20,
               p = .75)
  ) %>%
  mutate(dbinom = dbinom(x, n, p),
         pbinom = pbinom(x, n, p),
         reject_region = ifelse(pbinom < 0.005 | pbinom > 0.995,
                                "0.01",
                                ifelse(pbinom < 0.025 | pbinom > 0.975,
                                       "0.05",
                                       ifelse(pbinom < 0.05 | pbinom > .95,
                                              "0.10",
                                              "non-reject"))))

ggplot(data = DF,
       mapping = aes(x = x,
                     y = dbinom)) + 
  geom_bar(data = filter(DF, reject_region %in% "non-reject"),
           stat = "identity",
           color = "black",
           fill = "lightgray") + 
  geom_bar(data = filter(DF, !reject_region %in% "non-reject"),
           mapping = aes(fill = reject_region),
           stat = "identity",
           color = "black") +
  scale_fill_manual(values = rev(pallette_green)) + 
  labs(fill = "alpha") + 
  ylab("Probability") + 
  facet_grid(p ~ .)
```

In the one-sided test, $\alpha$ is placed in only one tail. The figures below depict examples of rejection regions for selected values of the Binomial distribution parameters. In each case, $\alpha$ is the area in the tail of the figure. It follows, then, that the decision rule for a lower tailed test is:

Reject $H_0$ when $x \leq Binomial(\alpha, n, \pi_0)$

For an upper tailed test, the decision rule is:

Reject $H_0$ when $x \geq Binomial(1 - \alpha, n, \pi_0)$

```{r, echo = FALSE, fig.path = 'figures/', fig.cap='The examples displayed use $n = 20$. For the top, middle, and bottom examples, $\\pi$ is set at 0.3, 0.5, and 0.75, respectively.'}
DF <- 
  bind_rows(
    data.frame(x = 0:20,
               n = 20,
               p = .3,
               side = "Left Tailed",
               stringsAsFactors = FALSE),
    data.frame(x = 0:20,
               n = 20,
               p = .5,
               side = "Left Tailed",
               stringsAsFactors = FALSE),
    data.frame(x = 0:20,
               n = 20,
               p = .75,
               side = "Left Tailed",
               stringsAsFactors = FALSE),
    data.frame(x = 0:20,
               n = 20,
               p = .3,
               side = "Right Tailed",
               stringsAsFactors = FALSE),
    data.frame(x = 0:20,
               n = 20,
               p = .5,
               side = "Right Tailed",
               stringsAsFactors = FALSE),
    data.frame(x = 0:20,
               n = 20,
               p = .75,
               side = "Right Tailed",
               stringsAsFactors = FALSE)
  ) %>%
  mutate(dbinom = dbinom(x, n, p),
         pbinom = pbinom(x, n, p),
         reject_region = ifelse(side == "Left Tailed",
                                ifelse(pbinom < 0.01,
                                       "0.01",
                                       ifelse(pbinom < 0.05,
                                              "0.05",
                                              ifelse(pbinom < 0.10,
                                                     "0.10",
                                                     "non-reject"))),
                                ifelse(pbinom > 0.99,
                                       "0.01",
                                       ifelse(pbinom > 0.95,
                                              "0.05",
                                              ifelse(pbinom > 0.90,
                                                     "0.10",
                                                     "non-reject")))))

ggplot(data = DF,
       mapping = aes(x = x,
                     y = dbinom)) + 
  geom_bar(data = filter(DF, reject_region %in% "non-reject"),
           stat = "identity",
           color = "black",
           fill = "lightgray") + 
  geom_bar(data = filter(DF, !reject_region %in% "non-reject"),
           mapping = aes(fill = reject_region),
           stat = "identity",
           color = "black") +
  scale_fill_manual(values = rev(pallette_green)) + 
  labs(fill = "alpha") + 
  ylab("Probability") + 
  facet_grid(p ~ side)
```

### Power

The derivations below make use of the following symbols:

* $x$: The observed frequency of experimental units exhibiting the trait of interest.
* $n$: The total number of experimental units.
* $\pi_0$: The proportion of the population that exhibits the trait of interest under the null hypothesis.
* $\pi_a$: The proportion of the population that exhibits the trait of interest under the alternative hypothesis.
* $\alpha$: The significance level.
* $\gamma(\pi)$: The power of the test for the parameter $\pi$.
* $Binomial(\alpha, n, \pi)$: A quantile of the Binomial distribution with a probability $\alpha$, and parameters $n$ and $\pi$.
* $C$: The critical region.

**Two Sided Test**

$$\begin{align}
\gamma(\pi_a) &= P_{\pi_a}(x \in C) \\
 &= P_{\pi_a}(Binomial(\alpha/2, n, \pi_0) \leq Binomial(\alpha / 2, n, \pi_a)) + \\
 & \ \ \ \ \ \ \ P_{\pi_a}(Binomial(1 - \alpha / 2, n, \pi_0) \geq Binomial(1 - \alpha / 2, n, \pi_a))
\end{align}$$

**Left Sided Test**

$$\begin{align}
\gamma(\pi_a) &= P_{\pi_a}(x \in C) \\
 &= P_{\pi_a}(Binomial(\alpha, n, \pi_0) \leq Binomial(\alpha, n, \pi_a)) 
\end{align}$$

**Right Sided Test**

$$\begin{align}
\gamma(\pi_a) &= P_{\pi_a}(x \in C) \\
 &= P_{\pi_a}(Binomial(1 - \alpha, n, \pi_0) \geq Binomial(1 - \alpha, n, \pi_a)) 
\end{align}$$

Since the Binomial distribution is discrete, the power curve has the interesting characteristic of not being monotonic.  It is sometimes described as having a "sawtooth" appearance. This behavior means that a larger sample size is not always preferred.  For example, in the following figure, a sample size of 10 has better power than a sample size of 12.

```{r, echo = FALSE, fig.path = 'figures/', fig.cap='Power for a Binomial test with $\\pi_0 = .15$ and $\\pi_a = 0.25$'}
BinomialPower <- 
  test_binomial(n = 2:30, 
                p0 = .15, 
                p1 = .25, 
                alpha= .05, 
                power = NULL,
                tail = 'right')

ggplot(data = BinomialPower,
       mapping = aes(x = n,
                     y = power)) + 
  geom_line(color = pallette_green[1]) + 
  ylab("Power") + 
  xlab("Sample Size")
```
