---
title: "It Can Be Shown"
author: "Benjamin Nutter"
header-includes:
  - \usepackage{color}
  - \usepackage{chngcntr}
  - \counterwithout{section}{chapter}
  - \renewcommand{\chaptername}{}
  - \renewcommand{\thechapter}{}
  - \usepackage{pifont}
  - \renewcommand\thefootnote{\ding{\numexpr171+\value{footnote}}}
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
documentclass: book
---
```{r, echo=FALSE}
library(ggplot2)
library(RColorBrewer)
greens.pal <- rev(brewer.pal(9, "Greens"))
```

The content of these pages is intended to make explicit the development of
mathematical and statistical theory while avoiding the nebulous "It can be shown."
As much as possible, the equations are developed step-by-step so that each 
step can be followed and understood.  

In instances where the changes are not obvious, explanations are provided in 
red.  Originally, these explanations were intended to be footnotes, but footnotes
are not well suited to the web page format.  For consistency, the PDF version
of these pages is produced in an identical format.  The PDF may be downloaded
from the [https://github.com/nutterb/ItCanBeShown](It Can Be Shown) GitHub 
repository.  

Additions, corrections, and suggestions are welcome on GitHub.

\part{Probability Distributions}
```{r, child="md_to_pdf/Bernoulli.Rmd"}
```

```{r, child="md_to_pdf/BinomialDistribution.Rmd"}
```

```{r, child="md_to_pdf/ChiSquareDistribution.Rmd"}
```

\chapter{ The Exponential Distribution}
\chapter{ The Gamma Distribution}
\chapter{ The Geometric Distribution}
\chapter{ The Hypergeometric Distribution}
\chapter{ The Multinomial Distribution}
\chapter{ The Normal Distribution}
\chapter{ The Poisson Distribution}
\chapter{ The Skew-Normal Distribution}
\chapter{ The Uniform Distribution}
\chapter{ The Weibull Distribution}

\part{Frequentist Hypothesis Testing}

\chapter{ Mantel-Haenszel Test}

\part{Supplemental Subjects}

\chapter{ Chebychev's Theorem}
\chapter{ Combinations}
\chapter{ The Correlation Coefficient}
\chapter{ Covariance}
\chapter{ Experimental Designs}
\chapter{ Moments and Moment Generating Functions}
\chapter{ Summation}
\chapter{ The Method of Transformations}
\chapter{ Variance Paramter}

\part{Non-Statistical Proofs}

\chapter{ The Binomial Theorem}
```{r, child="md_to_pdf/Functions.Rmd"}
```
\chapter{ The Geometric Series}
\chapter{ Integraion: Techniques and Theorems}
\chapter{ Logarithmic and Exponential Functions}
\chapter{ The Real Number System}


