---
title: "Notes on Statistical Theory"
author: "Benjamin Nutter"
header-includes:
  - \usepackage{color}
  - \usepackage{chngcntr}
  - \counterwithout{section}{chapter}
  - \renewcommand{\chaptername}{}
  - \renewcommand{\thechapter}{}
  - \usepackage{pifont}
  - \renewcommand\thefootnote{\ding{\numexpr171+\value{footnote}}}
output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
documentclass: book
---
```{r, echo=FALSE}
library(ggplot2)
library(RColorBrewer)
greens.pal <- rev(brewer.pal(9, "Greens"))
```

\part{Probability Distributions}
```{r, child="md_to_pdf/Bernoulli.Rmd"}
```

```{r, child="md_to_pdf/BinomialDistribution.Rmd"}
```

```{r, child="md_to_pdf/ChiSquareDistribution.Rmd"}
```

\chapter{ The Exponential Distribution}
\chapter{ The Gamma Distribution}
\chapter{ The Geometric Distribution}
\chapter{ The Hypergeometric Distribution}
\chapter{ The Multinomial Distribution}
\chapter{ The Normal Distribution}
\chapter{ The Poisson Distribution}
\chapter{ The Skew-Normal Distribution}
\chapter{ The Uniform Distribution}
\chapter{ The Weibull Distribution}

\part{Frequentist Hypothesis Testing}

\chapter{ Mantel-Haenszel Test}

\part{Supplemental Subjects}

\chapter{ Chebychev's Theorem}
\chapter{ Combinations}
\chapter{ The Correlation Coefficient}
\chapter{ Covariance}
\chapter{ Experimental Designs}
\chapter{ Moments and Moment Generating Functions}
\chapter{ Summation}
\chapter{ The Method of Transformations}
\chapter{ Variance Paramter}

\part{Non-Statistical Proofs}

\chapter{ The Binomial Theorem}
\chapter{ Functions}
\chapter{ The Geometric Series}
\chapter{ Integraion: Techniques and Theorems}
\chapter{ Logarithmic and Exponential Functions}
\chapter{ The Real Number System}


