# T-test

## Definitions and Terminology

## One-Sample T-test

The t-test is commonly used to look for evidence that the mean of a 
normally distributed random variable may differ from a hypothesized (or 
previously observed) value. 

### T-Statistic

The $t$-statistic is a standardized measure of the magnitude of difference between a sample's mean and some known, non-random constant.  It is similar to a $z$-statistic, but differs in that a $t$-statistic may be calculated without knowledge of the population variance.

### Definitions and Terminology

Let $\bar{x}$ be a sample mean from a sample with standard deviation $s$.  Let $\mu_0$ be a constant, and $s_\bar{x} = s/\sqrt{n}$ be the standard error of the parameter $\bar{x}$.  $t$ is defined:
$$t = \frac{\bar{x} - \mu_0}{s_\bar{x}} = \frac{\bar{x} - \mu_0}{\frac{s}{\sqrt{n}}}$$

### Hypotheses

The hypotheses for these test take the forms:

For a two-sided test:
$$
  \begin{align*}
  H_0: \mu &= \mu_0\\
  H_1: \mu &\neq \mu_0
  \end{align*}
$$

For a one-sided test:
$$
  \begin{align*}
  H_0: \mu &< \mu_0\\
  H_1: \mu &\geq \mu_0
  \end{align*}
$$

or 
$$
  \begin{align*}
  H_0: \mu &> \mu_0\\
  H_1: \mu &\leq \mu_0
  \end{align*}
$$

To compare a sample $(X_1, \ldots, X_n)$ against the hypothesized value, a 
T-statistic is calculated in the form:

$$T = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

Where $\bar{x}$ is the sample mean and $s$ is the sample standard deviation.

### Decision Rule

The decision to reject a null hypothesis is made when an observed T-value lies
in a critical region that suggests the probability of that observation is low.
We define the critical region as the upper bound we are willing to accept for
$\alpha$, the Type I Error.

In the two-sided test, $\alpha$ is shared equally in both tails.  The rejection
regions for the most common values of $\alpha$ are depicted in the figure below, 
with the sum of shaded areas on both sides equaling the corresponding $\alpha$.
It follows, then, that the decision rule is:

Reject $H_0$ when $T \leq t_{\alpha/2, n-1}$ or when $T \geq t_{1-\alpha/2, n-1}$.

By taking advantage of the symmetry of the T-distribution, we can simplify the
decision rule to:

Reject $H_0$ when $|T| \geq t_{1-\alpha/2, n-1}$

```{r, echo=FALSE, fig.height=3, fig.width=6, fig.path = 'figures/', fig.cap="The example displayed uses 25 degrees of freedom"}
RejectRegion <- 
  data.frame(x = seq(-5, 5, by = 0.01)) %>%
  mutate(y = dt(x, df = 24),
         sig = cut(pt(x, df = 24),
                   breaks = c(0, 0.005, 0.025, 0.05, 
                              0.95, 0.975, 0.995, 1),
                   labels = 1:7))
levels(RejectRegion[["sig"]]) = list("0.01" = c(1,7),
                                     "0.05" = c(2,6),
                                     "0.10" = c(3,5))

ggplot(data = RejectRegion,
       mapping = aes(x = x)) + 
  geom_line(mapping = aes(y = y)) + 
  geom_area(data = filter(RejectRegion, !is.na(sig) & x < 0),
            mapping = aes(y = y,
                          ymax = y,
                          fill = sig)) + 
  geom_area(data = filter(RejectRegion, !is.na(sig) & x > 0),
            mapping = aes(y = y,
                          ymax = y,
                          fill = sig)) + 
  scale_fill_manual(values = pallette_green) +  
  xlab("T-value") + 
  ylab("Probability") + 
  labs(fill="alpha")
```

In the one-sided test, $\alpha$ is placed in only one tail.  The rejection
regions for the most common values of $\alpha$ are depicted in the figure below.
In each case, $\alpha$ is the area in the tail of the figure.
It follows, then, that the decision rule for a lower tailed test is:

Reject $H_0$ when $T \leq t_{\alpha, n-1}$.

For an upper tailed test, the decision rule is:

Reject $H_0$ when $T \geq t_{1-\alpha, n-1}$.

Using the symmetry of the T-distribution, we can simplify the 
decision rule as:

Reject $H_0$ when $|T| \geq t_{1-\alpha, n-1}$.

```{r, echo=FALSE, fig.height=3, fig.width=6, fig.path = 'figures/', fig.cap="The example displayed uses 25 degrees of freedom"}
RejectRegion <- 
  expand.grid(x = seq(-5, 5, by = 0.01),
              tail = c("Lower Tailed", "Upper Tailed")) %>%
  mutate(y = dt(x, df = 24),
         sig = cut(pt(x, df = 24),
                   breaks = c(0, 0.005, 0.025, 0.05, 
                              0.95, 0.975, 0.995, 1),
                   labels = 1:7),
         sig = as.numeric(sig),
         sig = ifelse(tail %in% "Lower Tailed" & x > 0,
                      4, sig),
         sig = ifelse(tail %in% "Upper Tailed" & x < 0,
                      4, sig),
         sig = factor(sig))
levels(RejectRegion[["sig"]]) = list("0.01" = c(1,7),
                                     "0.05" = c(2,6),
                                     "0.10" = c(3,5))

ggplot(data = RejectRegion,
       mapping = aes(x = x)) + 
  geom_line(mapping = aes(y = y)) + 
  facet_grid(~ tail) + 
  geom_area(data = filter(RejectRegion, !is.na(sig) & x < 0),
            mapping = aes(y = y,
                          ymax = y,
                          fill = sig)) + 
  geom_area(data = filter(RejectRegion, !is.na(sig) & x > 0),
            mapping = aes(y = y,
                          ymax = y,
                          fill = sig)) + 
  scale_fill_manual(values = pallette_green) +  
  xlab("T-value") + 
  ylab("Probability") + 
  labs(fill="alpha")
```

The decision rule can also be written in terms of $\bar{x}$:

Reject $H_0$ when $\bar{x} \leq \mu_0 - t_\alpha \cdot s/\sqrt{n}$ or
$\bar{x} \geq \mu_0 + t_\alpha \cdot s/\sqrt{n}$.

This change can be justified by:

$$
\begin{align*}
|T| &\geq t_{1-\alpha, n-1}\\
\Big|\frac{\bar{x} - \mu_0}{s/\sqrt{n}}\Big| &\geq t_{1-\alpha, n-1} 
\end{align*}
$$



$$
\begin{align*}
-\Big(\frac{\bar{x} - \mu_0}{s/\sqrt{n}}\Big) &\geq t_{1-\alpha, n-1} &
    \frac{\bar{x} - \mu_0}{s/\sqrt{n}} &\geq t_{1-\alpha, n-1}\\
\bar{x} - \mu_0 &\leq - t_{1-\alpha, n-1} \cdot s/\sqrt{n} &
    \bar{x} - \mu_0 &\geq t_{1-\alpha, n-1} \cdot s/\sqrt{n}\\
\bar{x} &\leq \mu_0 - t_{1-\alpha, n-1} \cdot s/\sqrt{n} &
    \bar{x} &\geq \mu_0 + t_{1-\alpha, n-1} \cdot s/\sqrt{n} 
\end{align*}
$$

For a two-sided test, both the conditions apply.  The left side condition is used
for a left-tailed test, and the right side condition for a right-tailed test.

### Power

The derivations below make use of the following symbols:

* $\bar{x}$: The sample mean
* $s$: The sample standard deviation
* $n$: The sample size
* $\mu_0$: The value of population mean under the null hypothesis
* $\mu$: The value of the population mean under the alternative hypothesis.
* $\alpha$: The significance level
* $\gamma(\mu)$: The power of the test for the parameter $\mu$.
* $t_{\alpha, n-1}$: A quantile of the central t-distribution for a probability,
    $\alpha$ and $n-1$ degrees of freedom.
* $T$: A calculated value to be compared against a t-distribution.
* $C$: The critical region (rejection region) of the test.


**Two-Sided Test**

$$
\begin{align*}
\gamma(\mu) &= P_\mu(\bar{x} \in C)\\
&= P_\mu\big(\bar{x} \leq \mu_0 - t_{\alpha/2, n-1} \cdot s/\sqrt{n}\big) +
               P_\mu\big(\bar{x} \geq \mu_0 + t_{1-\alpha/2, n-1} \cdot s/\sqrt{n}\big)\\
&= P_\mu\big(\bar{x} - \mu \leq \mu_0 - \mu - t_{\alpha/2, n-1} \cdot s/\sqrt{n}\big) +
   P_\mu\big(\bar{x} - \mu \geq \mu_0 - \mu + t_{1-\alpha/2, n-1} \cdot s/\sqrt{n}\big)\\
&= P_\mu\Big(\frac{\bar{x} - \mu}{s/\sqrt{n}} \leq 
        \frac{\mu_0 - \mu - t_{\alpha/2, n-1} \cdot s/\sqrt{n}}{s/\sqrt{n}}\Big) +
   P_\mu\Big(\frac{\bar{x} - \mu}{s/\sqrt{n}} \geq 
        \frac{\mu_0 - \mu + t_{1-\alpha/2, n-1} \cdot s/\sqrt{n}}{s/\sqrt{n}}\Big)\\
&= P_\mu\Big(T \leq \frac{\mu_0 - \mu}{s/\sqrt{n}} - t_{\alpha/2, n-1}\Big) + 
   P_\mu\Big(T \geq \frac{\mu_0 - \mu}{s/\sqrt{n}} + t_{1-\alpha/2, n-1}\Big)\\
&= P_\mu\Big(T \leq -t_{\alpha/2, n-1} + \frac{\mu_0 - \mu}{s/\sqrt{n}}\Big) + 
   P_\mu\Big(T \geq t_{1-\alpha/2, n-1} + \frac{\mu_0 - \mu}{s/\sqrt{n}}\Big)\\
&= P_\mu\Big(T \leq -t_{\alpha/2, n-1} + \frac{\sqrt{n} \cdot (\mu_0 - \mu)}{s}\Big) + 
   P_\mu\Big(T \geq t_{1-\alpha/2, n-1} + \frac{\sqrt{n} \cdot (\mu_0 - \mu)}{s}\Big)
\end{align*}
$$

Both $t_{\alpha/2, n-1}$ and $t_{1-\alpha/2, n-1}$ have non-central T-distributions
with non-centrality parameter $\frac{\sqrt{n} (\mu_0 -\mu)}{s}$.
   

**One-Sided Test**

For convenience, the power for only the upper tailed test is derived here.  
Recall that the symmetry of the t-test allows us to use the decision rule:
Reject $H_0$ when $|T| \geq t_{1-\alpha}$.  Thus, where $T$ occurs in the 
derivation below, it may reasonably be replaced with $|T|$.

$$
\begin{align*}
\gamma(\mu)  &= P_\mu(\bar{x} \in C)\\
&= P_\mu\big(\bar{x} \geq \mu_0 + t_{1-\alpha, n-1} \cdot s / \sqrt{n}\big)\\
&= P_\mu\big(\bar{x} - \mu \geq \mu_0 - \mu + t_{1-\alpha, n-1} \cdot s / \sqrt{n}\big)\\
&= P_\mu\Big(\frac{\bar{x} - \mu}{s/\sqrt{n}} \geq 
    \frac{\mu_0 - \mu + t_{1-\alpha, n-1} \cdot s / \sqrt{n}}{s / \sqrt{n}}\Big)\\
&= P_\mu\Big(T \geq \frac{\mu_0 - \mu}{s/\sqrt{n}} + t_{1-\alpha, n-1} \Big)\\
&= P_\mu\Big(T \geq t_{1-\alpha, n-1} + \frac{\mu_0 - \mu}{s/\sqrt{n}}\Big)\\
&= P_\mu\Big(T \geq t_{1-\alpha, n-1} + \frac{\sqrt{n} \cdot (\mu_0 -\mu)}{s}\Big)
\end{align*}
$$

Where $t_{1-\alpha, n-1} + \frac{\sqrt{n} (\mu_0 -\mu)}{s}$ has a non-central 
t-distribution with non-centrality parameter $\frac{\sqrt{n} (\mu_0 -\mu)}{s}$

### Confidence Interval

The confidence interval for $\theta$ is written:
$$\bar{x} \pm t_{1-\alpha/2} \cdot \frac{s}{\sqrt{n}}$$

The value of the expression on the right is often referred to as the _margin of error_, and we will refer to this value as 
$$E = t_{1-\alpha/2} \cdot \frac{s}{\sqrt{n}}$$

## Two-Sample T-test

### T-Statistic

### Decision Rule

### Power

### Confidence Interval